# Introduction to R {#introduction}

## Getting Started

The simplest way to use R is to use it as if it were a calculator. For example, if we want to know what one plus one is, you may type:

```{r results='hide', message=FALSE, eval=FALSE}
  1 + 1
```

We can use any arithmetic operator, like addition, subtraction, multiplication, divison, exponentiation, and modulus operations:


```{r results='hide', message=FALSE, eval=FALSE}
  # addition
  1 + 1
  
  # subtraction
  6 - 4
  
  # multiplication 
  2 * 2 
  
  # division
  10 / 5
  10.2 / 5
  
  # integer divison
  15.2 %/% 5 
  15.7 %/% 5
  
  # modulus
  15.2 %% 5 
  15.8 %% 5
  
  # exponential
  2^3 
  
```

R also provides numerous built-in functions to use in calculations, such as natural logs, exponentiation, square root, absolute value:

```{r results='hide', message=FALSE, eval=FALSE}
  # natural log
  log(10)
  
  # exponentiation
  exp(2)
  
  # square root 
  sqrt(4)
  
  # absolute value
  abs(-4)
  
```
R includes extensive facilities for accessing documentation and searching for help. This is useful to get more information about a specific function. The `help()` function and `?` help operator in R provide access to the documentation pages for R functions. For example, to get help with the round() function, we submit the following code:

```{r results='hide', message=FALSE, eval=FALSE}
  #help() function
  help(round)

  #? operator  
  ?round()
```

## Variable Assignment 

A basic concept in programming is called a variable. A variable allows you to store a value (e.g. 10) or an object (e.g. a function description) in R. We can then further use the variable's name to access the value or the object that is stored within this variable.

For example, you can assign a value 10 to a variable `my_variable`:

```{r results='hide', message=FALSE, eval=FALSE}
  my_variable <- 10
```

To print out the value of the variable, you simply type the name of your variable:

```{r results='hide', message=FALSE, eval=FALSE}
  my_variable
```

A valid variable name consists of letters, numbers and the dot or underline characters. The variable name starts with a letter or the dot not followed by a number:

```{r results='hide', message=FALSE, eval=FALSE}
  # A variable can be operated on 
  my_variable + 1
  
  # And passed to a function
  sqrt(my_variable + 1)
  
  # To reassign a variable, just reassign in
  my_variable <- 3000

  # You can also operate on and reassign a variable to itself 
  my_variable <- my_variable + 1
```

You can broaden assignments beyond numbers:

```{r results='hide', message=FALSE, eval=FALSE}
  result <- sqrt(9)
  
  fruit_1 <- "apple"
  fruit_2 <- "banana"
  fruit_3 <- "cantaloupe"
```


## Finding Variables

To know all the variables currently available in the workspace we use the `ls()` function:

```{r results='hide', message=FALSE, eval=FALSE}
  print(ls())
```

The `ls()` function can also use patterns to match the variable names.

```{r results='hide', message=FALSE, eval=FALSE}
  # List the variables starting with the pattern "var".
  print(ls(pattern = "var"))   
```

## Deleting Variables

Variables can be deleted by using the `rm()` function. Below we delete the variable my_variable:

```{r results='hide', message=FALSE, eval=FALSE}
  my_variable <- 5
  rm(my_variable)
```

All the variables can be deleted by using the `rm()` and `ls()` function together:

```{r results='hide', message=FALSE, eval=FALSE}
  rm(list = ls())
```

Another common way to remove all variables in the R environment is to click on the little broom icon next to the button `Import Dataset` under the Environment tab. One can also go up to `Session` and do `Restart R` or `New Session` (the `Restart R` and `New session` options might come in handy for when programs crash).