# Data Structures

## Vectors

A vector is the simplest type of data structure in R. Vectors are particularly important as most of the functions you will write will work with vectors. Simply put, a vector is a sequence of data elements of the same basic type. 

We can construct a vector using the combine `(c)` function:

```{r prompt=TRUE, eval=FALSE}
  # Here is a vector containing three numeric values 3, 5 and 7:
  numbers <- c(3, 5, 7)

  #Here is a vector of logical values:
  logical <- c(TRUE, FALSE, TRUE, FALSE, FALSE) 
  
  #Here is a vector of character values:
  fruits <- c("apple", "oranges", "banana")
```

We can also change the value in a vector:

```{r prompt=TRUE, eval=FALSE}
  fruits[2] <- "strawberries"
```

Every vector has two key properties: its type and its length:

```{r prompt=TRUE, eval=FALSE}
  typeof(fruits)
  length(fruits)
```

You can combine vectors using the combine function as well:

```{r prompt=TRUE, eval=FALSE}
  results <- c(1, 2, 3)
  other_results <- c(4, 5, 6)
  combined_results <- c(results, other_results)
```

Vectors can be added together in an element-wise operation:

```{r prompt=TRUE, eval=FALSE}
  total_add <- results + other_results
  total_div <- results / other_results
  total_mod <- results %% other_results
```


Functions can be applied to each element of the vector. However, not all functions are vectorized (we will cover this later):

```{r prompt=TRUE, eval=FALSE}
  total_rou <- round(total_div, 1)

  names <- c("Simmons", "Race", "Healey", "LaBarr", "Villanes")
  names <- sort(names, decreasing = TRUE)
```

There are also functions for constructing useful types of vectors:

```{r prompt=TRUE, eval=FALSE}
  # Replicate - replicates a value X number of times 
  identity_vector <- rep(1,10)
  identity_vector
  
  # Sequence operator ":" From:To 
  sequence_vector <- 1:10
  
  # Sequence function - allows sequencing by a value
  sequence_by_two_vector <- seq(0,10, 2)
  sequence_by_two_vector
  
  # Sample function - take a random sample from a vector
  random_vector <- sample(sequence_by_two_vector,3)
```

Finally, a vector doesn't have to be made with consistent elements, but it will force them to one type:

```{r prompt=TRUE, eval=FALSE}
  values <- c("IAA", 1, "2021", 5)
  typeof(values)
```

As we mentioned earlier, the elements of a vector can only be of one type. If we want elements of different types... we will need a list.


## Lists

## Arrays

## Matrices

## Data Frames

## Working with different types
