# Introduction to R

## Getting Started

The simplest way to use R is to use it as if it were a calculator. For example, if we want to know what one plus one is, you may type:

```{r prompt=TRUE, eval=FALSE}
  1 + 1
```

We can use any arithmetic operator, like addition, subtraction, multiplication, division, divison, exponentiation, and modulus operations:


```{r prompt=TRUE, eval=FALSE}
  # addition
  1 + 1
  
  # subtraction
  6 - 4
  
  # multiplication 
  2 * 2 
  
  # division
  10 / 5
  10.2 / 5
  
  # integer divison
  15.2 %/% 5 
  15.7 %/% 5
  
  # modulus
  15.2 %% 5 
  15.8 %% 5
  
  # exponential
  2^3 
  
```

R also provides numerous built-in functions to use in calculations, such as natural logs, exponentiation, square root, absolute value:

```{r prompt=TRUE, eval=FALSE}
  # natural log
  log(10)
  
  # exponentiation
  exp(2)
  
  # square root 
  sqrt(4)
  
  # absolute value
  abs(-4)
  
```
R includes extensive facilities for accessing documentation and searching for help. This is useful to get more information about a specific function. The `help()` function and `?` help operator in R provide access to the documentation pages for R functions. For example, to get help with the round() function, we submit the following code:

```{r}
  #help() function
  help(round)

  #? operator  
  ?round()
```

## Variable Assignment

A basic concept in programming is called a variable. A variable allows you to store a value (e.g. 10) or an object (e.g. a function description) in R. We can then further use the variable's name to access the value or the object that is stored within this variable.

For example, you can assign a value 10 to a variable `my_variable`:

```{r prompt=TRUE, eval=FALSE}
  my_variable <- 10
```

To print out the value of the variable, you simply type the name of your variable

```{r prompt=TRUE, eval=TRUE}
  my_variable
```