# Programming Fundamentals

## Logicals

A logical is a binary representation of True and False:

```{r prompt=TRUE, eval=FALSE}
  a <- TRUE
  
  b <- FALSE
  
  typeof(b)
```

Logicals are used for evaluating comparisons, such as:

```{r prompt=TRUE, eval=FALSE}
  #equality
  # note the double equals == operator is a logical comparison, opposed to a single equal = being an assignment 
    
  2==2
  typeof(2==2)
    
  'cat' == 'dog'
    
  'cat' == 'cat'
    
  #Not equal operator 
  2!=2 
  
  # greater than/less than
  2>2
  
  2>=2
    
  2<1
    
  2<=1
    
  # Null Values 
  V1 <- 1
  V1[10] <- 10
    
  is.na(V1)
  !is.na(V1)
    
  # Contained within set 
  V1
    
  1 %in% V1
  10 %in% V1
  2 %in% V1
  
    
  c(1,2) %in% V1
    
  # If we want not in, then use the not ! operator around the entire statement 
  !(1 %in% V1)
    
  # Note, logicals can be used with other data types as well
  a <- c(1,2,3,4,5) 
  a <= 1
    
  b <- c(1,2,7,9,5)
  a == b
  a != b
    
  A <- matrix(1:10,2,5)
  B <- matrix(seq(1,20,2),2,5)
    
  A == B
  A != B
    
  identical(a,b)
  identical(A,B)
    
  all.equal(A,B)
```

## Conditionals

Conditionals are expressions that perform different computations or actions depending on whether a predefined boolean condition is TRUE or FALSE. Conditionals allow us to control the flow of execution. We use logicals to evaluate conditional statements. Conditionals include if, else, ifelse.

The syntax of if statement is:

```{r prompt=TRUE, eval=FALSE}
  if (test_expression) {
  statement
  }
```

If the test_expression is `TRUE`, the statement gets executed. But if it is `FALSE`, nothing happens. For example:

```{r prompt=TRUE, eval=FALSE}
  x <- 5
  if(x > 0){
  print("Positive number")
  }
```

If you will be executing a single statement after the conditional, you can withold the { }

```{r prompt=TRUE, eval=FALSE}
  x <- 5
  if(x > 0) print("Positive number")
```

If we want code to execute when the test_expression is `FALSE`, then we add an else statement:

```{r prompt=TRUE, eval=FALSE}
  if (test_expression) {
  statement
  } else {
  statement2
  }
```

For example:

```{r prompt=TRUE, eval=FALSE}
  x <- -5
  if(x > 0){
  print("Non-negative number")
  } else {
  print("Negative number")
  }
```

Note that you can use multiple logicals in the test_expression, such as:

* Or operator | 
* And operator &

For example:

```{r prompt=TRUE, eval=FALSE}
  mean_values <- 3.1
  std_dev_values <- 1.95
  
  if (mean_values>=0 & std_dev_values>=1) {
      print(paste("The mean", mean_values, "is above 0, and the standard deviation", std_dev_values, "is above 1"))
  }
```

Additionally, you can use the ifelse function to quickly assign a value based on a condition. The syntax for the ifelse function is `ifelse(test_expression, yes, no)`. For example:

```{r prompt=TRUE, eval=FALSE}
  std_dev_vector <- c(1.2,0.8,0.3,2.4)
  ifelse(std_dev_vector>=1, "above one", "below one")
```

## Functions

## Loops

## Vectorization

```{r prompt=TRUE, eval=FALSE}

```